---
title: "Severe Weather Events Investigation"
author: "Johan Jordaan"
date: "16 March 2016"
subtitle: "... a short investigation of the economic and population health impact of severe weather events in the United States ..."
output: 
  html_document:
    keep_md: true
---

## Synopsis
This investigation aimed to answer the following two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The data used comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

## Assumptions 
I made some assumption in the investigation in order to answer the questions. The following list enumerates these assumptions:

* I assumed the American definition for a billion, which is a thousand million.
* I assumed that 'harmful with respect to population health' is measured by number of injuries and fatalities. 
* I assumed that 'have the greatest economic consequences' is measured by the damage to propery and crops and not include harmful efects on population healt.
* I have excluded any kind of normilisation of values to account for inflation.
* The analsysis has been done on the aggregate data in the database stretching from 1950 to 2011. 

## Data Processing
```{r warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r load, cache=TRUE}
data <- read.csv("repdata-data-StormData.csv.bz2")
```

### Data Quality
The data quality was quite high in my oppinion with only the PROPDMGEXP and CROPDMGEXP variables needing some kind of cleanup.

####Event Type Data Quality
The EVTYPE variable contained no NA values and required no cleanup.

```{r }
sum(is.na(data$EVTYPE))
```


####Injury/Fatality Data Quality
The FATALITIES and INJURIES variables contained no NA or negative values and required no cleanup.

```{r }
sum(is.na(data$FATALITIES))
sum(data$FATALITIES<0)
sum(is.na(data$INJURIES))
sum(data$INJURIES<0)
```


####Property Damage Data Quality
The PROPDMG variable contained no NA or negative values and required no cleanup.

```{r }
sum(is.na(data$PROPDMG))
sum(data$PROPDMG<0)
```

The PROPDMGEXP variable should have only contained K(Thousands), M(Millions) and B(Billions) values. Since it contained other values this series nedded to be cleaned.

```{r }
table(data$PROPDMGEXP)
```

####Crop Damage Data Quality
The CROPDMG variable contained no NA or negative values and required no cleanup.

```{r }
sum(is.na(data$CROPDMG))
sum(data$CROPDMG<0)
```

The CROPDMGEXP variable should have only contained K(Thousands), M(Millions) and B(Billions) values. Since it contained other values this series nedded to be cleaned.

```{r }
table(data$CROPDMGEXP)
```

### Data Cleanup and Enhancement

I replaced all the values in the PROPDMGEXP and CROPDMGEXP variables that fell outside of the valid range (K,M,B) with the most populous value, namely K.I further compensated for possible capitlisation discrepancies.

```{r cleanup}
m <- function(exp) {
  ifelse(exp %in% c("B","b"),1000000*1000,ifelse(exp %in% c("M","m"),1000000,1000))  
}

data <- data %>% mutate(ABSPROPDMG = PROPDMG * m(PROPDMGEXP), ABSCROPDMG = CROPDMG * m(CROPDMGEXP))

```

### Data Aggregation

In the aggregate I normalised the damage amount to billions by deviding the total amounts by one billion. I applied this normalisation here to reduce rounding effects. I further transformed the data to 'tidy data' to make plotting and further analysis easier.

```{r aggregate}
top <- 10

data <- data %>%
        group_by(EVTYPE) %>%
        summarize( TOT_FATALITIES=sum(FATALITIES)
                  ,TOT_INJURIES=sum(INJURIES)
                  ,TOTAL_HEALTH=TOT_FATALITIES+TOT_INJURIES
                  ,TOT_PROPDMG=sum(ABSPROPDMG)/100000000
                  ,TOT_CROPDMG=sum(ABSCROPDMG)/100000000
                  ,TOTAL_DAMAGE=TOT_PROPDMG+TOT_CROPDMG
                  )

health <- data %>% 
          filter(TOTAL_HEALTH>0) %>% 
          select(EVTYPE,TOT_FATALITIES,TOT_INJURIES,TOTAL_HEALTH) %>%
          arrange(desc(TOTAL_HEALTH)) %>%
          top_n(top,TOTAL_HEALTH) %>%
          gather(TYPE,VALUE,TOT_FATALITIES:TOTAL_HEALTH) %>% 
          mutate(EVTYPE=factor(EVTYPE),TYPE=factor(TYPE,labels=c("Total","Fatalities","Injuries")))

damage <- data %>% 
          filter(TOTAL_DAMAGE>0) %>% 
          select(EVTYPE,TOT_PROPDMG,TOT_CROPDMG,TOTAL_DAMAGE) %>%
          arrange(desc(TOTAL_DAMAGE)) %>%
          top_n(top,TOTAL_DAMAGE) %>%
          gather(TYPE,VALUE,TOT_PROPDMG,TOT_PROPDMG:TOTAL_DAMAGE) %>% 
          mutate(EVTYPE=factor(EVTYPE),TYPE=factor(TYPE,labels=c("Total","Crop Damage","Property Damage")))
```

Note that I only selected the top ```r top``` TOTAL_DAMAGE and TOTAL_HEALTH elements in the list. 

```{r}
health
damage
```

## Exploratory Graphs
### Health Impact by Event Type
```{r graph_health}
ggplot(health,aes(reorder(EVTYPE,-VALUE),VALUE,color=TYPE,group=TYPE)) + 
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Event Type") +
  ylab("Event Impact - Individuals") +
  ggtitle("Population health impact of the worst weather events in the United States")
```

### Damage by Event Type
```{r graph_damage}
ggplot(damage,aes(reorder(EVTYPE,-VALUE),VALUE,color=TYPE,group=TYPE)) + 
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Event Type") +
  ylab("Event Impact - Billion $") +
  ggtitle("Economic impact of the worst weather events in the United States")
```

## Results

The goal of the investigation was to answer these two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The answers to these two questions given the assumptions and the data analysis are:

1. Accross the United States, Hurricanes are the most harmfull with respect to population health.
2. Accross the United States, Floods have the greatest economic consequences.


## Further Work

The investigation was constraigned by the requirements and time limits. In doing this investigation the following open questions ocurred to me and might be interesting engigh for furure analisies following on from this work:

1. What is the progression over time for the events with the highest impact. We know that hurricanes has the higest impact on population health, but has this always been the case?
2. Would the results of this study be diffrent if the values are infation/market adjusted?
3. Is there any events that in aggregate terms has less impact but has a higher average and accure more frequently. This might indicate 'low hanging fruit' for event preperation.
4. Is there a way to quantify injuries and fatalities sothat they can be added to the economic loss of property and crop in order to get a better economic impact.
5. Is the impact of events localised in terms of economic and population healt impact?

These are but a few of the questions that popped into my head while doing this investigation. I am looking forward to answering some of the above questions and in ansering them generating a whole herd of new questions.






