---
title: "Severe Weather Events Investigation"
author: "Johan Jordaan"
date: "16 March 2016"
output: 
  html_document:
    keep_md: true
---

## Synopsis
This investigation aimed to answer the following two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The data used comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

## Assumptions 
I assumed the American definition for a billion, which is a thousand million.

## Data Processing
```{r warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r load, cache=TRUE}
data <- read.csv("repdata-data-StormData.csv.bz2")
```

### Data Quality
The data quality was quite high in my oppinion with only the PROPDMGEXP and CROPDMGEXP variables needing some kind of cleanup.

####Event Type Data Quality
The EVTYPE variable contained no NA values and required no cleanup.

```{r }
sum(is.na(data$EVTYPE))
```


####Injury/Fatality Data Quality
The FATALITIES and INJURIES variables contained no NA values and required no cleanup.

```{r }
sum(is.na(data$FATALITIES))
sum(is.na(data$INJURIES))
```


####Property Damage Data Quality
```{r }
sum(is.na(data$PROPDMG))
```

The PROPDMG variable contained no NA values and required no cleanup.

```{r }
table(data$PROPDMGEXP)
```

The PROPDMGEXP variable should have only contained K(Thousands), M(Millions) and B(Billions) values. Since it contained other values this series nedded to be cleaned.

####Crop Damage Data Quality
```{r }
sum(is.na(data$CROPDMG))
```

The CROPDMG variable contained no NA values and required no cleanup.

```{r }
table(data$CROPDMGEXP)
```

The CROPDMGEXP variable should have only contained K(Thousands), M(Millions) and B(Billions) values. Since it contained other values this series nedded to be cleaned.

### Data Cleanup and Enhancement

I replaced all the values in the PROPDMGEXP and CROPDMGEXP variables that fell outside of the valid range (K,M,B) with the most populous value, namely K.I further compensated for possible capitlisation discrepancies.

```{r cleanup}
m <- function(exp) {
  ifelse(exp %in% c("B","b"),1000000*1000,ifelse(exp %in% c("M","m"),1000000,1000))  
}

data <- data %>% mutate(ABSPROPDMG = PROPDMG * m(PROPDMGEXP), ABSCROPDMG = CROPDMG * m(CROPDMGEXP))

```

### Data Aggregation

In the aggregate I normalised the damage amount to billions by deviding the total amounts by one billion. I applied this normalisation here to reduce rounding effects. I further transformed the data to 'tidy data' to make plotting and further analysis easier.

```{r aggregate}
top <- 10

data <- data %>%
        group_by(EVTYPE) %>%
        summarize( TOT_FATALITIES=sum(FATALITIES)
                  ,TOT_INJURIES=sum(INJURIES)
                  ,TOTAL_HEALTH=TOT_FATALITIES+TOT_INJURIES
                  ,TOT_PROPDMG=sum(ABSPROPDMG)/100000000
                  ,TOT_CROPDMG=sum(ABSCROPDMG)/100000000
                  ,TOTAL_DAMAGE=TOT_PROPDMG+TOT_CROPDMG
                  )

health <- data %>% 
          filter(TOTAL_HEALTH>0) %>% 
          select(EVTYPE,TOT_FATALITIES,TOT_INJURIES,TOTAL_HEALTH) %>%
          arrange(desc(TOTAL_HEALTH)) %>%
          top_n(top,TOTAL_HEALTH) %>%
          gather(TYPE,VALUE,TOT_FATALITIES:TOTAL_HEALTH) %>% 
          mutate(EVTYPE=factor(EVTYPE),TYPE=factor(TYPE,labels=c("Total","Fatalities","Injuries")))

damage <- data %>% 
          filter(TOTAL_DAMAGE>0) %>% 
          select(EVTYPE,TOT_PROPDMG,TOT_CROPDMG,TOTAL_DAMAGE) %>%
          arrange(desc(TOTAL_DAMAGE)) %>%
          top_n(top,TOTAL_DAMAGE) %>%
          gather(TYPE,VALUE,TOT_PROPDMG,TOT_PROPDMG:TOTAL_DAMAGE) %>% 
          mutate(EVTYPE=factor(EVTYPE),TYPE=factor(TYPE,labels=c("Total","Crop Damage","Property Damage")))
```

Note that I only selected the top ```r top``` TOTAL_DAMAGE and TOTAL_HEALTH elements in the list. 

```{r}
health
damage
```

## Exploratory Graphs
### Health Impact by Event Type
```{r graph_health}
ggplot(health,aes(reorder(EVTYPE,-VALUE),VALUE,color=TYPE,group=TYPE)) + 
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Event Type") +
  ylab("Event Impact - Individuals") 
```

### Damage by Event Type
```{r graph_damage}
ggplot(damage,aes(reorder(EVTYPE,-VALUE),VALUE,color=TYPE,group=TYPE)) + 
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Event Type") +
  ylab("Event Impact - Billion $") 
```

## Results

The goal of the investigation was to answer these two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The answers to these two questions given the assumptions and the data analysis are:

1. Accross the United States, Hurricanes are the most harmfull with respect to population health.
2. Accross the United States, Floods have the greatest economic consequences.


## Further Work




