---
title: "Severe Weather Events Investigation"
author: "Johan Jordaan"
date: "16 March 2016"
output: 
  html_document:
    keep_md: true
---

## Synopsis
Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

Across the United States, which types of events (as indicated in the ğ™´ğš…ğšƒğšˆğ™¿ğ™´ variable) are most harmful with respect to population health?
Across the United States, which types of events have the greatest economic consequences?

## Data Processing
```{r warning=FALSE,message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```

```{r load, cache=TRUE}
data <- read.csv("repdata-data-StormData.csv.bz2")
```

```{r }
dim(data)
```

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the ğšŒğšŠğšŒğš‘ğš = ğšƒğšğš„ğ™´ option for certain code chunks.

## Results
At least one but not more than 3 figures



